---
title: "Methods"
number-sections: true
---

```{r}
#| label: methods-setup
#| include: false
library(dissertation)
library(data.table)
data(assessmentData)
```

## Overview {#sec-methods-overview}

This chapter describes the data, measures, and analytic methods used to address the research questions. All analyses are implemented as reproducible R functions within the `dissertation` R package, ensuring complete transparency and replicability.

## Data Source {#sec-data-source}

The dataset comprises `r format(nrow(assessmentData), big.mark = ",")` longitudinal assessment records for `r format(length(unique(assessmentData$ID)), big.mark = ",")` students across grades `r paste(sort(unique(assessmentData$GRADE)), collapse = ", ")` in `r paste(unique(assessmentData$CONTENT_AREA), collapse = " and ")` spanning `r length(unique(assessmentData$YEAR))` academic years (`r paste(sort(unique(assessmentData$YEAR)), collapse = ", ")`).

## Sample Description {#sec-sample}

```{r}
#| label: tbl-sample
#| tbl-cap: "Sample Size by Content Area and Year"
#| echo: false

sample_summary <- assessmentData[, .(
    N_Students = uniqueN(ID),
    N_Observations = .N
), keyby = .(CONTENT_AREA, YEAR)]

knitr::kable(sample_summary, format.args = list(big.mark = ","))
```

The sample includes students from `r length(unique(assessmentData$DISTRICT_NUMBER))` districts and `r length(unique(assessmentData$SCHOOL_NUMBER))` schools. @tbl-sample presents the sample composition by content area and year.

## Measures {#sec-measures}

### Scale Scores

Student achievement is measured using scale scores in Reading and Mathematics. Scale scores range from `r min(assessmentData$SCALE_SCORE)` to `r max(assessmentData$SCALE_SCORE)` across the dataset.

### Achievement Levels

Scale scores are classified into performance levels using grade-specific cut points. The levels in the dataset are: `r paste(sort(unique(assessmentData$ACHIEVEMENT_LEVEL)), collapse = ", ")`. Students at the Proficient or Advanced level are considered to have met grade-level expectations.

### Organizational Variables

Records include district identifiers (`DISTRICT_NUMBER`) and school identifiers (`SCHOOL_NUMBER`), enabling analyses of between-district and between-school variation.

## Analytic Approach {#sec-analytic-approach}

The analyses are organized to address each research question:

**RQ1 — Achievement Trends**: Longitudinal achievement patterns are examined by computing mean scores and proficiency rates by year and content area using the `analyzeAssessment("achievement_trends")` function.

**RQ2 — District Variation**: Between-district comparisons use `analyzeAssessment("district_comparison")` to examine variation in mean scores, proficiency rates, and sample sizes across the `r length(unique(assessmentData$DISTRICT_NUMBER))` districts in the dataset.

**RQ3 — Cohort Progress**: Cohort tracking via `analyzeAssessment("cohort_progress")` examines how students with longitudinal records progress through grade levels over time.

Additionally, score distribution analyses using `analyzeAssessment("score_distribution")` provide percentile breakpoints (10th, 25th, 50th, 75th, 90th) to characterize the full shape of the achievement distribution.

## Reproducibility {#sec-reproducibility}

All analyses are implemented in the `dissertation` R package and can be replicated by installing the package and calling the analysis functions directly:

```{r}
#| label: reproducibility-demo
#| eval: false
library(dissertation)
summarizeAssessment(content_area = "READING", group_by = "GRADE")
analyzeAssessment("achievement_trends", content_area = "MATHEMATICS")
```

The complete analysis code, data, and documentation are available via the package's GitHub repository and REST API.
