---
title: "Results"
number-sections: true
---

```{r}
#| label: results-setup
#| include: false
library(dissertation)
library(data.table)
data(assessmentData)
```

## Overview {#sec-results-overview}

This chapter presents the findings organized by research question. All results are generated programmatically from the `dissertation` R package analysis functions, ensuring that the narrative is always synchronized with the underlying data.

## RQ1: Achievement Trends {#sec-rq1}

```{r}
#| label: achievement-trends
#| echo: false
trends <- analyzeAssessment("achievement_trends")
```

Achievement trends across `r length(unique(trends$results$YEAR))` academic years reveal the following patterns:

```{r}
#| label: tbl-achievement-trends
#| tbl-cap: "Mean Score and Proficiency Rate by Year and Content Area"
#| echo: false
knitr::kable(trends$results, digits = 1)
```

`r trends$interpretation`

### Reading Achievement Patterns

```{r}
#| label: reading-trends
#| echo: false
reading_trends <- analyzeAssessment("achievement_trends", content_area = "READING")
```

`r reading_trends$interpretation`

### Mathematics Achievement Patterns

```{r}
#| label: math-trends
#| echo: false
math_trends <- analyzeAssessment("achievement_trends", content_area = "MATHEMATICS")
```

`r math_trends$interpretation`

## RQ2: District Variation {#sec-rq2}

### Between-District Achievement Comparisons

```{r}
#| label: district-comparison
#| echo: false
district <- analyzeAssessment("district_comparison")
```

```{r}
#| label: tbl-district
#| tbl-cap: "Achievement by District"
#| echo: false
knitr::kable(district$results, digits = 1)
```

`r district$interpretation`

### Achievement Summaries by Grade

```{r}
#| label: grade-summary
#| echo: false
grade_summary <- summarizeAssessment(group_by = "GRADE")
```

```{r}
#| label: tbl-grade-summary
#| tbl-cap: "Achievement Summary by Grade"
#| echo: false
knitr::kable(grade_summary$summary, digits = 1)
```

## RQ3: Cohort Progress {#sec-rq3}

```{r}
#| label: cohort-progress
#| echo: false
cohort <- analyzeAssessment("cohort_progress")
```

```{r}
#| label: tbl-cohort
#| tbl-cap: "Cohort Progress: Students with Longitudinal Records"
#| echo: false
knitr::kable(cohort$results, digits = 1)
```

`r cohort$interpretation`

## Score Distribution Properties {#sec-distribution}

```{r}
#| label: distribution
#| echo: false
dist <- analyzeAssessment("score_distribution")
```

```{r}
#| label: tbl-distribution
#| tbl-cap: "Score Distribution by Content Area and Grade"
#| echo: false
knitr::kable(dist$results, digits = 1)
```

`r dist$interpretation`

## Summary of Findings {#sec-results-summary}

The results reveal several key patterns across the three research questions. Achievement trends show the trajectory of proficiency rates and mean scores over the years examined. District comparisons reveal variation in outcomes that warrants further investigation. The cohort analysis demonstrates how student achievement trajectories develop over the longitudinal span of the dataset, while score distribution analyses reveal the full shape of the achievement continuum beyond mean-level summaries.
