---
title: "Data Dictionary"
---

## assessmentData

The primary dataset contains longitudinal student assessment records across multiple years, grades, districts, and schools.

```{r}
#| label: setup
#| include: false
library(dissertation)
library(data.table)
data(assessmentData)
```

### Variables

| Variable | Type | Description | Example Values |
|----------|------|-------------|----------------|
| CONTENT_AREA | Character | Subject area | READING, MATHEMATICS |
| YEAR | Character | Academic year (YYYY_YYYY format) | 2022_2023, 2023_2024, 2024_2025 |
| GRADE | Character | Grade level | 3, 4, 5, 6, 7, 8 |
| ID | Character | Unique student identifier | Anonymized student IDs |
| SCALE_SCORE | Numeric | Assessment scale score | Continuous numeric scores |
| ACHIEVEMENT_LEVEL | Character | Performance level classification | Beginning, Developing, Proficient, Advanced |
| DISTRICT_NUMBER | Integer | District identifier | Numeric district codes |
| SCHOOL_NUMBER | Integer | School identifier | Numeric school codes |

### Summary Statistics

```{r}
#| label: tbl-data-summary
#| tbl-cap: "Dataset Dimensions"
#| echo: false
cat("Rows:", format(nrow(assessmentData), big.mark = ","), "\n")
cat("Columns:", ncol(assessmentData), "\n")
cat("Students:", format(length(unique(assessmentData$ID)), big.mark = ","), "\n")
cat("Years:", paste(sort(unique(assessmentData$YEAR)), collapse = ", "), "\n")
cat("Grades:", paste(sort(unique(assessmentData$GRADE)), collapse = ", "), "\n")
cat("Districts:", length(unique(assessmentData$DISTRICT_NUMBER)), "\n")
cat("Schools:", length(unique(assessmentData$SCHOOL_NUMBER)), "\n")
```

### Key: ID x CONTENT_AREA x YEAR

The dataset is keyed on (ID, CONTENT_AREA, YEAR), meaning each row represents one student's assessment in one subject for one year.
